---
title: 'Regression: Canadian Oil and Gas Stock Price Prediction using Company Balance
  Sheets'
output:
  html_document:
    df_print: paged
---

```{r, include = FALSE}
# importing packages
library(olsrr) # for model selection
library(GGally) # for checking for higher order terms
library(lmtest) # for the Breusch-Pagan test
library(mctest) # for the VIF test
library(MASS) # for the Box-Cox transformation
# suppresses scientific notation in outputs
options(scipen = 999)
# increases the maximum rows for outputs
options(max.print = 999999)
```

### Retrieving the Dataset

```{r}
# importing the csv file
data = read.csv("stock_price_prediction_oil_gas.csv")
# removing the Date and X columns from the dataset as we do not need them for our model
data = subset(data, select = -c(Date, X))
```

### Model Selection

```{r}
# creating the full base model with all of the variables in the dataset
model = lm(ClosingStockPrice~., data = data)
```

```{r}
# stepwise model selection
stepmod = ols_step_both_p(model, pent = 0.1, prem = 0.3, details = FALSE)
summary(stepmod$model)
```

```{r}
# backwards model selection
backmodel = ols_step_backward_p(model, prem = 0.3, details = FALSE)
summary(backmodel$model)
```

```{r}
# forward model selection
formodel = ols_step_forward_p(model, penter = 0.1, details = FALSE)
summary(formodel$model)
```

```{r}
# choosing consistantly present variables from all of the model selection methods
model = lm(ClosingStockPrice~Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets+Inventory, data = data)
summary(model)
```

### Checking for Higher Order Terms

Now, we can check for any terms with a higher order relationship with ClosingStockPrice.

```{r}
# checking for higher order terms
ggpairs(data[, c("ClosingStockPrice", "Ordinary.Shares.Number", "Total.Equity.Gross.Minority.Interest", 
    "Non.Current.Deferred.Taxes.Liabilities", "Total.Liabilities.Net.Minority.Interest", 
    "Net.Tangible.Assets", "Inventory")])
```

It appears there are no higher order terms present in the model.

### Checking for Interaction Terms

Now, we can check for any interaction terms between the variables in the model.

```{r}
# checking for interaction terms
interactmodel = lm(ClosingStockPrice~(Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets+Inventory)^2, data = data)
summary(interactmodel)
```

```{r}
# removing insignificant interaction terms
model = lm(ClosingStockPrice~Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets+Inventory+Ordinary.Shares.Number:Inventory+Total.Equity.Gross.Minority.Interest:Inventory+Non.Current.Deferred.Taxes.Liabilities:Total.Liabilities.Net.Minority.Interest+Total.Liabilities.Net.Minority.Interest:Inventory+Net.Tangible.Assets:Inventory, data = data)
summary(model)
```

```{r}
# removing insignificant interaction terms
model = lm(ClosingStockPrice~Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets+Inventory+Ordinary.Shares.Number:Inventory, data = data)
summary(model)
```

```{r}
# removing insignificant interaction terms
model = lm(ClosingStockPrice~Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets+Inventory, data = data)
summary(model)
```

### Condition Checking

Now, we must check the 6 core regression assumptions that are necissary to be met, for a sufficient regression model.

Linearity Assumption:

```{r}
# plotting the residual plot of the model
plot(model, which = 1)
```

It appears that the linearity assumption is not met, as the plot is not consistantly linear.

Normality Assumption:

```{r}
# plotting the residual histogram
hist(residuals(model), main = "Residual Histogram", xlab = "Residuals")
```

```{r}
# plotting the Q-Q normality plot
plot(model, which = 2)
```

```{r}
# conducting the Shapiro-Wilk normality test
shapiro.test(residuals(model))
```

$H_0: $ Normality of Residuals is met.

$H_A: $ Normality of Residuals is not met.

It appears that the normality of the residuals assumption is not met, as the histogram is not a bell-shape, the Q-Q normality plot is not following the diagonal line and we reject the null hypothesis in the Shapiro-Wilk normality test.

Equal Variance Assumption:

```{r}
# plotting the residual plot for the model
plot(model, which = 1)
```

```{r}
# plotting the scale-location plot
plot(model, which = 3)
```

```{r}
# conducting the Breusch-Pagan test
bptest(model)
```

$H_0: $The Equal Variance Assumption is met.

$H_A: $The Equal Variance Assumption is not met.

It appears that the equal variance assumption is not met, as the residual plot and scale location plot do not have equal variance for all x values and we reject the null hypothesis in the Breusch-Pagan test.

Independence Assumption:

Independence is assumed by default.

Multicollinearity Assumption:

```{r}
# conducting the vif test
imcdiag(model, method = "VIF")
```

It appears that the multicollinearity assumption is not met, as there is collinearity present between 3 variables.

Outliers:

```{r}
# plotting the residual-leverage plot
plot(model, which = 5)
```

```{r}
# plotting the cooks distance values
plot(model, pch = 18, col = "red", which = c(4))
```

It appears that the outlier assumption is met, as there are no cooks distance values present above 0.5.

### Box-Cox Transformation

Due to the fact that there are multiple assumptions that have not been met by our model, we can conduct a box-cox tranformation in an attempt to see if it can rectify this issue.

```{r}
# getting the best lambda value for the box-cox transformation
bc = boxcox(model, lambda = seq(-10, 10), data = data)
```

```{r}
# getting the best lambda value for the box-cox transformation
best.lambda = bc$x[which(bc$y == max(bc$y))]
# creating the box-cox model
model = lm(formula = (ClosingStockPrice^best.lambda - 1)/best.lambda ~Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets+Inventory, data = data)
summary(model)
```

```{r}
# removing the inventory variable as it is insignificant now
model = lm(formula = (ClosingStockPrice^best.lambda - 1)/best.lambda ~Ordinary.Shares.Number+Total.Equity.Gross.Minority.Interest+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets, data = data)
summary(model)
```

### Condition Checking

Now, we must check the 6 core regression assumptions that are necissary to be met on the box-cox model.

Linearity Assumption:

```{r}
# plotting the residual plot of the model
plot(model, which = 1)
```

It appears that the linearity assumption is now met.

Normality Assumption:

```{r}
# plotting the residual histogram
hist(residuals(model), main = "Residual Histogram", xlab = "Residuals")
```

```{r}
# plotting the Q-Q normality plot
plot(model, which = 2)
```

```{r}
# conducting the Shapiro-Wilk normality test
shapiro.test(residuals(model))
```

$H_0: $ Normality of Residuals is met.

$H_A: $ Normality of Residuals is not met.

It appears that the normality of the residuals assumption is not met, as the histogram is not a bell-shape, the Q-Q normality plot is not following the diagonal line and we reject the null hypothesis in the Shapiro-Wilk normality test.

Equal Variance Assumption:

```{r}
# plotting the residual plot of the model
plot(model, which = 1)
```

```{r}
# plotting the scale-location plot
plot(model, which = 3)
```

```{r}
# conducting the Breusch-Pagan test
bptest(model)
```

$H_0: $The Equal Variance Assumption is met.

$H_A: $The Equal Variance Assumption is not met.

It appears that the equal variance assumption is not met, as we reject the null hypothesis in the Breusch-Pagan test.

Independence Assumption:

Independence is assumed by default.

Multicollinearity Assumption:

```{r}
# conducting the vif test
imcdiag(model, method = "VIF")
```

It appears that the multicollinearity assumption is not met, as there is is collinearity present between 2 variables.

Outliers:

```{r}
# plotting the residual-leverage plot
plot(model, which = 5)
```

```{r}
# plotting the cooks distances
plot(model, pch = 18, col = "red", which = c(4))
```

It appears that the outlier assumption is met, as all cooks distances are below 0.5.

### Removing the collinear variable

```{r}
model = lm(formula = (ClosingStockPrice^best.lambda - 1)/best.lambda ~Ordinary.Shares.Number+Non.Current.Deferred.Taxes.Liabilities+Total.Liabilities.Net.Minority.Interest+Net.Tangible.Assets, data = data)
summary(model)
```

```{r}
# removing the insignificant variables
model = lm(formula = (ClosingStockPrice^best.lambda - 1)/best.lambda ~Ordinary.Shares.Number+Net.Tangible.Assets, data = data)
summary(model)
```

### Condition Checking

Now, we must check the 6 core assumptions for the final model.

Linearity Assumption:

```{r}
# plotting the residual plot of the model
plot(model, which = 1)
```

It appears that the linearity assumption is not met, as the residual plot is not linear across all x values.

Normality Assumption:

```{r}
# plotting the residual histogram
hist(residuals(model), main = "Residual Histogram", xlab = "Residuals")
```

```{r}
# plotting the Q-Q normality plot
plot(model, which = 2)
```

```{r}
# conducting the Shapiro-Wilk normality test
shapiro.test(residuals(model))
```

$H_0: $ Normality of Residuals is met.

$H_A: $ Normality of Residuals is not met.

It appears that the normality of the residuals assumption is not met, as the histogram is not a bell-shape, the Q-Q normality plot is not following the diagonal line and we reject the null hypothesis in the Shapiro-Wilk normality test.

Equal Variance Assumption:

```{r}
# plotting the residual plot of the model
plot(model, which = 1)
```

```{r}
# plotting the scale-location plot
plot(model, which = 3)
```

```{r}
# conducting the Breusch-Pagan test
bptest(model)
```

$H_0: $The Equal Variance Assumption is met.

$H_A: $The Equal Variance Assumption is not met.

It appears that the equal variance assumption is not met, as the residual and scale-location plots are not equally varied for all x values, and we reject the null hypothesis in the Breusch-Pagan test.

Independence Assumption:

Independence is assumed by default.

Multicollinearity Assumption:

```{r}
# conducting the vif test
imcdiag(model, method = "VIF")
```

It appears that the multicollinearity assumption is now met.

Outliers:

```{r}
# plotting the residual-leverage plot
plot(model, which = 5)
```

```{r}
# plotting the cooks distance values
plot(model, pch = 18, col = "red", which = c(4))
```

It appears that the outlier assumption is met.

### Results

In summarizing the project, we encountered challenges in constructing a robust regression model. These challenges included a scarcity of strong variable predictors and difficulties in meeting various regression model assumptions, both before and after applying a Box-Cox transformation. Notably, our findings indicate that company balance sheets lack efficacy as predictors for the closing stock prices of Canadian oil and gas companies. This limitation arises from the fact that most balance sheet values exhibit an insignificant relationship with the closing stock price. Additionally, those few balance sheet values that do demonstrate a significant relationship fail to satisfy fundamental regression requirements such as the equal variance and normality of residuals assumptions, rendering the model unsuitable for predictive purposes. These insights underscore the importance of exploring factors beyond financial statement values when evaluating potential investments in the oil and gas sector. Investors are encouraged to delve deeper into a company's mission statement, social media activity, management history, and other non-financial aspects to make more informed investing decisions.

### Conclusion

In conclusion, it appears that balance sheets may not be a reliable predictor for the closing stock prices of Canadian oil and gas companies. This is likely due to the influence of extraneous factors beyond the numerical values presented in financial statements that significantly impact the closing stock prices.

### References

Yfinance. PyPI. (2023). https://pypi.org/project/yfinance/
